---
title: "Car Crash Project"
author: "Ethan Schultz"
date: "2024-11-26"
output: html_document
---

Reading in the data and preparing predictor variables for analysis and visualization
```{r}
library(readxl)
library(ggplot2)
data <- read_xlsx("/Users/ethanschultz/Downloads/Big_Data_Files (1).xlsx", sheet = "Car Crash")
data$CrashType <- as.numeric(factor(data$CrashType, levels = c("A", "B", "C", "D", "E", "F", "G")))
data$CrashType <- factor(data$CrashType)
data$ViolCat <- factor(data$ViolCat)
```

Initial Data Exploration

```{r}
# Count occurrences of each class in Severity
table(data$Severity)

```

Made a function for plotting to easily see the number of crashes in the different categories

```{r}
create_barplot <- function(data, var_name) {
  var_summary <- table(data[[var_name]])
  
  ggplot(data = as.data.frame(var_summary), aes(x = Var1, y = Freq)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    labs(
      title = paste("Number of Crashes by", var_name),
      x = var_name,
      y = "Count"
    ) +
    theme_minimal()
}
```

```{r}
severity <- data[data$Severity == 1, ]
```

```{r}
head(severity)
```


```{r}

severity$CrashTypes <- factor(severity$CrashType,
                         levels = c(1, 2, 3, 4, 5, 6, 7),
                         labels = c("Head-on", "Sideswipe", "Rear End", 
                                    "Broadside", "Hit Object", "Overturned", 
                                    "Vehicle/Pedestrian"))

ggplot(severity, aes(x = CrashTypes, fill = factor(Severity))) +
  geom_bar(position = "dodge") +
  labs(title = "Counts of Crash Type by Severity", 
       x = "Crash Type", 
       y = "Count", 
       fill = "Severity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}
ggplot(severity, aes(x = CrashType, fill = Severity)) +
  geom_bar(position = "dodge") +
  labs(title = "Counts of Crash Type by Severity", 
       x = "Crash Type", 
       y = "Count")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
create_barplot(data,"CrashType")
```

```{r}
create_barplot(data,"ViolCat")
```

```{r}

severity$ViolCats <- factor(severity$ViolCat,
                         levels = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11),
                         labels = c("DUI/BUI", "Unsafe Speed", "Following too Closely", 
                                    "Improper Passing", "Unsafe Lane Chnage", "Improper Turning", 
                                    "Automobile Right of Way","Pedestrian Right of Way","Pedestrian Violation","Traffic Signals and Signs", "Fell Asleep"))

ggplot(severity, aes(x = ViolCats, fill = factor(Severity))) +
  geom_bar(position = "dodge") +
  labs(title = "Counts of Violation Categories by Severity", 
       x = "Violation Type", 
       y = "Count")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r}
create_barplot(data,"Weekday")
```

```{r}
create_barplot(data,"Severity")
```

```{r}
create_barplot(data,"ClearWeather")
```

```{r}
create_barplot(severity,"Highway")
```

```{r}
create_barplot(data,"Month")
```


After visually exploring the data first model was a logistic model to see how each predictor affected the severity of car crashes

Logistic model to predict severity
```{r}
model <- glm(Severity ~ ViolCat + ClearWeather + CrashType + Highway + Daylight + Highway*CrashType ,family = binomial ,data=data)
summary(model)
```


Necessary Packages for next model
```{r}
suppressWarnings(RNGversion("3.5.3"))
library(caret)
library(klaR)
library(gains)
library(pROC)
library(dplyr)
```


Next model was a Naive Bayes

Naive Bayes

Addressing data imbalance challenges by oversampling
```{r}
library(ROSE)

set.seed(1)
over <- ovun.sample(Severity ~ ., data = data1, method = "both", N = nrow(data1))$data

table(over$Severity)
head(over)

```

```{r}
data1 <-data[, !(names(data) %in% c("ID", "County", "City"))]

over$Severity <- as.factor(over$Severity)
set.seed(1)
myIndex<- createDataPartition(over$Severity, p=0.6, list=FALSE)
trainSet <- over[myIndex,]
validationSet <- over[-myIndex,]
myCtrl <- trainControl(method="cv", number=10)
set.seed(1)
nb_fit <- train(Severity ~., data = trainSet, method = "nb", trControl = myCtrl, tuneGrid = data.frame(fL = 0, usekernel = FALSE, adjust = 1))
nb_fit
nb_class <- predict(nb_fit, newdata = validationSet)
confusionMatrix(nb_class, validationSet$Severity, positive = '1')
```


Also tried undersampeling for comparison
```{r}
under <- ovun.sample(Severity~., data=data1, method = "under", N = 15836)$data
table(under$Severity)
```

```{r}

under$Severity <- as.factor(under$Severity)
set.seed(1)
myIndex<- createDataPartition(under$Severity, p=0.6, list=FALSE)
trainSet <- under[myIndex,]
validationSet <- under[-myIndex,]
myCtrl <- trainControl(method="cv", number=10)
set.seed(1)
nb_fit <- train(Severity ~., data = trainSet, method = "nb", trControl = myCtrl, tuneGrid = data.frame(fL = 0, usekernel = FALSE, adjust = 1))
nb_fit
nb_class <- predict(nb_fit, newdata = validationSet)
confusionMatrix(nb_class, validationSet$Severity, positive = '1')
```

Running into crazy amount of warnings when running because of the class imbalance in the data set there is significantly non-severe cases than severe cases in the data set. Still a lot of warnings but much less after I did the tuneGrid = ...




Association Rule Analysis to see common combinations of patters with different predictors
```{r}
library(arules)
library(arulesViz)

data2 <- data[, c("Severity", "ViolCat", "ClearWeather", "Highway", "Daylight")]

# Convert variables to factors if they're not already
data2$Severity <- as.factor(data2$Severity)
data2$ViolCat <- as.factor(data2$ViolCat)
data2$ClearWeather <- as.factor(data2$ClearWeather)
data2$CrashType <- as.factor(data2$CrashType)
data2$Highway <- as.factor(data2$Highway)
data2$Daylight <- as.factor(data2$Daylight)

transactions <- as(data2, "transactions")
rules <- apriori(transactions, parameter = list(support = 0.1, confidence = 0.5, minlen = 2))
srules <- sort(rules, by="lift", decreasing = TRUE)
inspect(srules)
plot(rules)
```


```{r}
data3 <- data[, c("Severity", "ClearWeather", "CrashType", "Highway", "Daylight")]

# Convert variables to factors if they're not already
data3$Severity <- as.factor(data3$Severity)
data3$ClearWeather <- as.factor(data3$ClearWeather)
data3$CrashType <- as.factor(data3$CrashType)
data3$Highway <- as.factor(data3$Highway)
data3$Daylight <- as.factor(data3$Daylight)

transactions <- as(data3, "transactions")
rules <- apriori(transactions, parameter = list(support = 0.1, confidence = 0.05, minlen = 2))
srules <- sort(rules, by="lift", decreasing = TRUE)
inspect(srules)
plot(rules)

rules_severity <- subset(rules, rhs %pin% "Severity")

rules_severity <- subset(rules_severity, confidence > 0.05)

inspect(rules_severity)

plot(rules_severity, method = "graph", engine = "htmlwidget")

```

